Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
wordsmtihing, formatting, subsectino headers

@@

This directed graph represents the states of a
hidden Markov model in the ``Plan7'' format defined by the HMMER
software package.
%% Only 7 of the possible 9 state transitions are allowed, which 
%% simplifies the dynamic programming.
Each insertion state~($I$) represents a potential mutation that 
inserts an amino acid;
each deletion state~($D$) represents a potental mutation that
deletes an amino acid;
and each match state~($M$) represents either
an unmutated amino acid or a potential mutation that
substitutes one amino acid for another.
A~match state or an insertion state includes a table that assigns a
nonzero probability to each amino acid.
(The figure shows tables for the first two states only.)
A~triplet of an insertion state, deletion state, 
and match state formas a \emph{node}, which corresponds to a column in
the alignment used to train the model.
(Each node is shown in a dashed oval.)

\caption{A hidden Markov model trained from a four-column alignment}

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

@@

\subsection{MRFy Implementation}

@@

@@

\subsection{Ugly code}

@@

>>>>

This model, trained from a four-column alignment,
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.

\caption{A hidden Markov model}

@@

@@

\begin{figure*}
\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\]

\centerline{$a'_{s} = - \log a_{s} \nonumber$,
$e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}$,
and 
$V_j^{\prime M}(i) = - V_j^{M}(i)$}
\caption{Viterbi's equations, in original and negated forms}
\label{viterbi}
\end{figure*}

@@

%\subsection{MRFy Implementation}

@@

\subsection{Higher-order functions}

@@

@@

\subsection{Awkward Haskell}
